A semantics of core AUTOSAR

### 2016-08-28 paper/paper.tex
Paper: read and made minor changes to the last part

The transition relation defined in Appendix~\ref{sec:Rules} is non-deterministic; i.e., it may very well be that for some process $p$,
$$
\begin{array}{c}
	p \red{l_1} p_1 \hspace{10mm}\text{and}\hspace{10mm} p \red{l_2} p_2
\end{array}
$$
are both possible. If it also holds that
$$
\begin{array}{c}
	p_1 \red{l_2} q \hspace{10mm}\text{and}\hspace{10mm} p_2 \red{l_1} q
\end{array}
$$
are derivable, we note that process $p$ may evolve into state $q$ either via an $l_1$ step followed by $l_2$, or by first taking the $l_2$ step and then $l_1$. Such freedom of choice is commonly taken as the definition of \emph{parallelism} in the process calculus literature --- if the sequential ordering of steps $l_1$ and $l_2$ has no significance, they might just as well be considered to occur in parallel.

@@

The AUTOSAR specification contains a very subtle sensitivity to this kind of non-deterministism that is worth examplifying. Consider a queued data element $a$ together with a runnable $b$ set up to trigger on data reception events for $a$. A scenario where two values $v_1$ and $v_2$ are sent on a port connected to $a$ may lead to the following trace:
$$
\begin{array}{c}
  {\Tqelem a 2 \Teps} \Opar
  {\Trunnable b 0 \Tidle 0} \vspace{0.6ex} \\

@@

  {\Tqelem a 2 {\Tcons{v_1}{v_2}}} \Opar
  {\Trunnable b 0 \Tidle 1} \,\Opar  \\
  {\Trinst b \Tvoid \Teps {\Code}}
\end{array}
$$
Here there is only one $b$ instance created and none is pending, even though $a$ correctly contains the two values received. The reason is that by swapping the order of the first instantiation and the second data reception steps, the system is put in a state where its inability to count pending instantiations becomes fatal.

@@

@@

Unqueued data elements that are read before first written should return the error code {\bf neverReceived} \cite[ch.~5.6.10]{AR:RTE}. Since no exceptions to this rule are mentioned in the specification, our intepretation is that this holds even when a data element possesses an \emph{initValue} attribute  (initial {\bf delem} state in Section~\ref{sec:Sem}). However, as this renders any such initial value assignments entirely redundant, an alternative interpretation would be to create a term
$$
  \Tdelem {i.p.e} \Tfalse v
$$
for each unqueued data element $i.p.e$, where $v$ is the \emph{initValue} of $e$ if it exists, and $\TneverReceived$ otherwise.

@@

The specification restricts client-server connections such that a client may not be connected to multiple servers, for the purpose of avoiding an operation call to be handled by more than one server runnable \cite[ch.~4.2.3]{AR:RTE}. However, this restriction does not explicitly forbid multiple \emph{OperationInvokedEvent}s refering to the same client-server operation to trigger separate runnables. Our formulation of client-server communication is based on the assumption that this is just an oversight, as allowing multiple events on a server operation would have the same problem with ambiguous response values as a would one-to-many client-server connection pattern. %It is not easy to see what could constitute an alternative interpretation without introducing some programmer option for selecting among server results.

@@

Activation of runnables due to \emph{DataReceivedEvent}s is faithfully captured in the formal semantics for unqueued as well as queued data elements (c.f.~axioms (17) and (29) with \cite[ch.~7.5.1.6]{AR:SWC}). But for unqueued elements AUTOSAR also provides an invalidation operation, whose net effect is very similar to that of a data write (see axiom (21)). Whether data invalidation should count as a reception event is not clear from the specification; we have chosen not to include it as a conservative conjecture. Revising this choice would not be hard, though: all that is necessary is to add an axiom that mimics (17) but matches on {\sc inv} rather than {\sc wr} labels.

A similar specification ambiguity concerns \emph{DataReceivedEvent}s in the presence of a full receiver buffer. We have chosen to block these events from activating connected runnables (axiom (30)), although the opposite choice could also be justified on grounds that activation of a reciever must be even more pressing when its buffer is overflowing. Expressing this formally would amount to simply deleting axiom (30) together with the last precondition of axiom (29).

@@

We claim that the formalized AUTOSAR semantics of this report covers a substantial core of the Software Component Template and Specification of RTE \cite{AR:SWC, AR:RTE}. But even these two documents alone add up to almost 2000 pages of specification text in total, covering a wide range of features on many levels of abstraction. We have therefore found it necessary to further limit our scope; both by leaving out parts that have very little to do with the concurrency and communication aspects that are our primary interest, and by ignoring parts that to some extent duplicate the features that we are already covering. We discuss the chosen omissions in more detail below.

@@

The only atomic software component type we include on our formalization is the \emph{ApplicationSwComponentType}. Other atomic component types describe dynamic behaviors that are either just minor variants of the formalized component type (\emph{SensorActuatorSwComponentType}, \emph{ComplexDeviceDriverSwComponentType}, \emph{ServiceSwComponentType}, \emph{ServiceProxySwComponentType}), or trivial in comparison (\emph{NvBlockSwComponentType}, \emph{EcuAbstractionSwComponentType}, \emph{ParameterSwComponentType}).

>>>>

The transition relation defined in Appendix~\ref{sec:Rules} is non-deterministic; i.e., it may very well be that for some process~$p$,
$$
\begin{array}{c}
	p \red{l_1} p_1 \hspace{10mm}\text{and}\hspace{10mm} p \red{l_2} p_2
\end{array}
$$
are both possible. If it also holds that
$$
\begin{array}{c}
	p_1 \red{l_2} q \hspace{10mm}\text{and}\hspace{10mm} p_2 \red{l_1} q
\end{array}
$$
are derivable, we note that process $p$ may evolve into state $q$ either via an $l_1$ step followed by $l_2$, or by first taking the $l_2$ step and then $l_1$. Such freedom of choice is commonly taken as the definition of \emph{parallelism} in the process calculus literature --- if the sequential ordering of steps $l_1$ and $l_2$ has no significance, they might just as well be considered to occur in parallel.

@@

The AUTOSAR specification contains a very subtle sensitivity to this kind of non-deterministism that is worth examplifying. Consider a queued data element $a$ together with a runnable $b$ set up to trigger on data reception events for $a$.
%**TODO: The explanation is rather long so it may be good to indicate in a few words what it will show already at start: "... lead to the runnable triggering once or twice"
A scenario where two values $v_1$ and $v_2$ are sent on a port connected to $a$ may lead to the following trace:
%
TODO: The i.p.e in the transition label is not explained.
$$
\begin{array}{c}
  {\Tqelem a 2 \Teps} \Opar
  {\Trunnable b 0 \Tidle 0} \vspace{0.6ex} \\

@@

  {\Tqelem a 2 {\Tcons{v_1}{v_2}}} \Opar
  {\Trunnable b 0 \Tidle 1} \,\Opar  \\
  {\Trinst b \Tvoid \Teps {\Code}}
\end{array}
$$
Here there is only one $b$ instance created and none is pending, even though $a$ correctly contains the two values received. The reason is that by swapping the order of the first instantiation and the second data reception steps, the system is put in a state where its inability to count pending instantiations becomes fatal.
%*TODO: Perhaps abbreviate the terms to make the difference more clear: let Q(vs) = \Tqelem a 2 {vs}; R(n) = \Trunnable b 0 \Tidle n; RI = \Trinst b \Tvoid \Teps {\Code}

%   Q(\Teps) \Opar R(0)  ->*  Q(\Tcons{v_1}{v_2}}) \Opar R(1) \Opar RI
% and
%   Q(\Teps) \Opar R(0)  ->*  Q(\Tcons{v_1}{v_2}}) \Opar R(2) \Opar RI \Opar RI

@@

%**TODO: The end of the previous section also seems to be a clarification proposal

@@

Unqueued data elements that are read before first written should return the error code {\bf neverReceived} \cite[ch.~5.6.10]{AR:RTE}. Since no exceptions to this rule are mentioned in the specification, our intepretation is that this holds even when a data element possesses an \emph{initValue} attribute  (initial {\bf delem} state in Section~\ref{sec:Sem}). However, as this renders any such initial value assignments entirely redundant, an alternative interpretation would be to create a term
$$
  \Tdelem {i.p.e} \Tfalse v
$$
for each unqueued data element $i.p.e$, where $v$ is the \emph{initValue} of $e$ if it exists, and $\TneverReceived$ otherwise.
%*TODO: perhaps replace i.p.e by a?

@@

The specification restricts client-server connections such that a client may not be connected to multiple servers, for the purpose of avoiding an operation call to be handled by more than one server runnable \cite[ch.~4.2.3]{AR:RTE}. However, this restriction does not explicitly forbid multiple \emph{Operation\-InvokedEvent}s refering to the same client-server operation to trigger separate runnables. Our formulation of client-server communication is based on the assumption that this is just an oversight, as allowing multiple events on a server operation would have the same problem with ambiguous response values as a would one-to-many client-server connection pattern. %It is not easy to see what could constitute an alternative interpretation without introducing some programmer option for selecting among server results.

@@

Activation of runnables due to \emph{DataReceived\-Event}s is faithfully captured in the formal semantics for unqueued as well as queued data elements (c.f.~axioms (17) and (29) with \cite[ch.~7.5.1.6]{AR:SWC}). But for unqueued elements AUTOSAR also provides an invalidation operation, whose net effect is very similar to that of a data write (see axiom (21)). Whether data invalidation should count as a reception event is not clear from the specification; we have chosen not to include it as a conservative conjecture. Revising this choice would not be hard, though: all that is necessary is to add an axiom that mimics (17) but matches on {\sc inv} rather than {\sc wr} labels.

A similar specification ambiguity concerns \emph{DataReceived\-Event}s in the presence of a full receiver buffer. We have chosen to block these events from activating connected runnables (axiom (30)), although the opposite choice could also be justified on grounds that activation of a reciever must be even more pressing when its buffer is overflowing. Expressing this formally would amount to simply deleting axiom (30) together with the last precondition of axiom (29).

@@

We claim that the formalized AUTOSAR semantics of this report covers a substantial core of the Software Component Template and Specification of RTE \cite{AR:SWC, AR:RTE}. But these two documents alone add up to almost 2000 pages of specification text in total, covering a wide range of features on many levels of abstraction. We have therefore found it necessary to further limit our scope; both by leaving out parts that have very little to do with the concurrency and communication aspects that are our primary interest, and by ignoring parts that to some extent duplicate the features that we are already covering. We discuss the chosen omissions in more detail below.

@@

%*TODO: Perhaps avoid repeating the common suffix SwComponentType (with explanation).
The only atomic software component type we include on our formalization is the \emph{ApplicationSwComponentType}. Other atomic component types describe dynamic behaviors that are either just minor variants of the formalized component type (\emph{SensorActuatorSwComponentType}, \emph{ComplexDeviceDriverSwComponentType}, \emph{ServiceSwComponentType}, \emph{ServiceProxySwComponentType}), or trivial in comparison (\emph{NvBlockSwComponentType}, \emph{EcuAbstractionSwComponentType}, \emph{ParameterSwComponentType}).
